---
title: "Practical: Data I/O"
author: "BaselRBootcamp April 2018"
output: html_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6, echo = TRUE, eval = TRUE)
```

### Slides

Here a link to the lecture slides for this session: <a href="https://therbootcamp.github.io/BaselRBootcamp_2018April/_sessions/D1S3_DataIO/Data_IO.html"><b>LINK</b></a>

### Overview

In this practical you'll learn how to read and save data. By the end of this practical you will know how to:

1. Identify the location of a file (on your hard drive)
2. Read in data of various types
3. Use R's file connections
4. Scrape the internet (advanced)

### Functions

Here are the main read-in functions:

| Function| Description|
|:------|:--------|
|     `read_csv()`| Read flat csv file|
|     `read_sas()`| Read SAS file|
|     `read_sav()`| Read SPSS file|
| `readRDS()`| Read RDS file |
| `file(...,'r'), readLines`| Read from file connection |

Here are the main export functions:

| Function| Description|
|:------|:--------|
|     `write_csv()`| Write flat csv file|
|     `write_sas()`| Write SAS file|
|     `write_sav()`| Write SPSS file|
| `saveRDS()`| Save RDS file |
| `file(...,'w'), readLines`| Write to file connection |


## Tasks

This tutorial begins with the titanic data set, which contains records of 1313 passengers of their name (`Name`), their passenger class (`PClass`), their age (`Age`), their sex (`Sex`), whether they survived (`Survived`), and a numeric coding of their sex (`SexCode`). 

Later you will be working with a randomly generated artificial data set, which will help demonstrate the relationship of file size and speed. 

```{r, eval = T, echo = F, include = FALSE}
# load a package
library(tidyverse)
library(haven)
```


### Identify the (file)path

1. Every read or write function requires that you provide a file path or source (e.g., an URL) specifying the location of the file. Ways to facilitate finding the right file path is to either set the working directory (see `?setwd()`) or, even better, to use projects. You will learn about projects. For now we will try to work with the full file path, that is beginning from your hard drive or user account. Fortunately, this isn't difficult either, when using RStudio's auto-complete functionality. To find a file path simply write to quotation marks, i.e., `""`, and press *<Tab>* with the cursor being in-between the quotation marks. This will open an auto-complete dialog that allows you to level-by-level search through your folder structure and to identify the correct file path. The auto-complete will always begin at the current working directory (see ?`getwd()`), which will usually be the level of your user account. Another way to begin at this location is to begin by writing `~/`. Try it out write `"~/"` into an R script, place the cursor after the slash (`/`), and press *<Tab>*. Now, try to find the location of each of the titanic data sets on your hard drive and assign the path (which is a character string) to an object, e.g., `titanic_csv_path` (titanic_path <- "~/folder_1/folder_2/.../titanic.csv").      

### Read to `tibble`

2. OK, let's read some data sets. First you need to load the packages `readr` and `haven` using `library()`, i.e., `library(readr)` and `library(haven)`.

3. Now, read in the three titanic data files "titanic.csv", "titanic.sav",  and "titanic.sas7bdat" using `read_csv()`, `read_sas()`, and `read_sav()` by passing on the respective file paths (i.e., `data/filename.csv`) and assigning the functions' outcomes to `df_csv`, `df_sas`, and `df_sav`, respectively. Inspect each of the imported objects using `print()` and `str()` and `typeof`. 

```{r, eval=T, echo=T}
df_csv <- read_csv('data/titanic.csv')
df_sas <- read_sas('data/titanic.sas7bdat')
df_sav <- read_sav('data/titanic.sav')
print(df_csv) ; print(df_sas) ; print(df_sav)
str(df_csv) ; str(df_sas) ; str(df_sav)
```

4. Verify that the three data sets are identical. To do this use the *is-equal-to* operator `==` introduced in the last tutorial and `all()`. 

```{r, eval=T, echo=T}
all(df_csv == df_sas)
all(df_csv == df_sav)
all(df_sas == df_sav)
```

5. Received an error? This is because the data sets contain `NA`'s, that is, missing values. One way to see this is via the summary function `summary()`. Try it out for each of the data sets.

```{r, eval=T, echo=T}
summary(df_csv)
summary(df_sas)
summary(df_sav)
```

6. Fortunately, `NA`'s aren't much of a problem. When you read `?all` (i.e., the help file), you can see that `all()` has a second argument called `na.rm = FALSE`. Using this argument, you can tell R to ignore `NA`'s. The default is set to `FALSE`, thus, this feature is deactivated. To activate it you must pass `na.rm = T` to `all()` (as the second argument). Try it. Are the data frames now equal? 

```{r, eval=T, echo=T}
all(df_csv == df_sas, na.rm = T)
all(df_csv == df_sav, na.rm = T)
all(df_sas == df_sav, na.rm = T)
```

### Write to disk

Every read function in R has a corresponding write function. In this section, you will write data to the three previous formats and R's own `.RData` and `.RDS` formats. Before turning to writing, however, you want to take care of a aspect associated with older read and write functions: `rownames`.

7. As you can see, the first column in, e.g., `df_csv` contains row numbers. Older write functions by default add a row number column to the data, when it writes data to disk. Newer functions, however, do not exhibit this behavior. Overall, `rownames` have come a bit out of fashion. Thus, we want to eliminate the first column from `df_csv` using deletion by negative indices. Negative indices in subsetting functions such as `[]` or `[[]]` means *omit* rather than *select*. E.g., `c(1, 2, 3)[-2]` returns the vector without the second value. Try to apply this to `df_csv` and assign the result to `df_csv_no_rownames`. You are dealing with a data frame. Should you use single `[]` or double `[[]]` brackets? Try both.

```{r, eval=T, echo=T}
df_csv_no_rownames = df_csv[-1]
#df_csv_no_rownames = df_csv[[-1]]
```

8. Once you removed the first column, write the reduced data frame to disk using `write_csv()`. Try using the exact same file path. Does R give you a warning? 

9. Now, let's try some other formats using the other data set. First, read in `my_data.csv` (which you downloaded in the beginning of this practical) as assign to `my_data`. Then, write `my_data` to disk using `write_sas`, `write_sav`, and also `saveRDS`. Just as `write_csv`, each of these functions take the data frame as the first argument and the file path as the second. To obtain the latter adapt the file path by hand to make sure that the file path has the correct file ending, i.e., `.sas7bdat`, `.sav`, and `.RDS`, respectively. While you write the individual data sets pay attention to the speed of execution. Which one was fastest? If you didn't make out a difference, try again.

```{r, eval=T, echo=T}
my_data = read_csv("data_files/my_data.csv")
write_sas(my_data,"data_files/my_data.sas7bdat")
write_sav(my_data,"data_files/my_data.sav")
saveRDS(my_data,"data_files/my_data.RDS")
```

10. You may have noticed that the `saveRDS` function was the slowest. This is because `.RDS` results compresses the data much more heavily and, thus, also resulrs in the smallest file sizes. To verify this use the `file.info()`-function on each of the four files, that is, on their file paths. Look out for the `size` element, which gives the size in `bytes`.

```{r, eval=T, echo=T}
file.info("data_files/my_data.csv")
file.info("data_files/my_data.sas7bdat")
file.info("data_files/my_data.sav")
file.info("data_files/my_data.RDS")
```

### Read from url

Most read functions are not limited to reading from the hard drive. Often you can, for instance, also directly read from url. 

11. Read the the *titanic* data set from this link *https://raw.githubusercontent.com/therbootcamp/therbootcamp.github.io/master/_sessions/_data/titanic.csv* using `read_csv()`.

```{r, eval=T, echo=T}
my_data = read_csv("https://raw.githubusercontent.com/therbootcamp/therbootcamp.github.io/master/_sessions/_data/titanic.csv")
```

### File connections

The most basic way to handle files is via file connections. In this more advanced section we take a look at how this works. 

12.  The first step of working with file connections is to establish a connections using `file()`. The main arguments to `file()` are a location (a file path, URL, etc.) and a mode indicator, e.g., `r` for `Open for reading in text mode`. Try opening a connection to the `titanic.csv` data set. To do this, you must assign the output of `file()` to an object, which then becomes the connection (e.g., `con = file("my_path","r")`). If you want to use the URL from above use `url()` instead of `file()`.

```{r, eval=T, echo=T}
con = file('data_files/titanic.csv', 'r')
```

13. Now that the connection has been established, you can read the file using `readLines()`. `readLines()` iterates through the file line by line and returns each line as a character string. Try reading the file using `readLines()` and store the output in an object. When done close the connection using `close(my_con)` and inspect the data.  

```{r, eval=T, echo=T}
lin = readLines(con)
close(con)
```

14. As you can see the read-in data looks much messier than when using `read_csv()`. This is because `read_csv()` really does a lot of editing and interpreting for you. If you're up for the challenge you can try to process the raw data. One approach is to first split each line using the `str_split` (package `stringr` from the `tidyverse`) function using the comma `,` as the `pattern` argument. Applying `str_split` will return a `list` of `vector`s, where each vector represents one row of the future data frame. Next, bind the rows to a matrix using `do.call(rbind, my_list)`. The only thing then left for you to do is transform the matrix to a data frame (using `as_data_frame` or, better, `as_tibble`) and to take care of column types and names.  

```{r, eval=T, echo=T}
require(stringi)
spl <- stri_split_fixed(lin,',')
mat <- do.call(rbind,spl)
df  <- as.tibble(mat)
# take care of types
```

### Scraping the internet

15. Finally on to something completely new. As mentioned, connections can also be established to data located outside one's own hard drive, such as, for instance, the world wide web. Specifically, using `url()` you can establish a connection to a webpage, which you then can use to retrieve information from the webpage. One package that conveniently streamlines extracting information from webpages based is the `rvest` package. To illustrate how easy it can be to scrape (aka extract) information from an webpage using R and `rvest`, run the following code. It downloads and parses the Milestones table from R's Wikipedia page.

```{r, eval=T, echo=T}

# load package
install.packages('rvest', repos = "https://stat.ethz.ch/CRAN/")
library(rvest)
library(tibble)
library(magrittr)

# get html
url = 'https://en.wikipedia.org/wiki/R_(programming_language)' 
page = read_html(url)

# get table
# use XPath from inspect page (e.g., in Chrome)
table = page %>% html_node(xpath = '//*[@id="mw-content-text"]/div/table[2]') %>% html_table()

# create tibble
as.tibble(table)
```


# Additional reading

- For more details on all steps of data analysis check out Hadley Wickham's [R for Data Science](http://r4ds.had.co.nz/).

- For more advanced content on objects check out Hadley Wickham's [Advanced R](http://adv-r.had.co.nz/).

- For more on pirates and data analysis check out the respective chapters in YaRrr! The Pirate's Guide to R [YaRrr! Chapter Link](https://bookdown.org/ndphillips/YaRrr/htests.html)




